Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-05 Manuscript/manuscript.Rmd
manuscript edits

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. Likewise, emerging research has shown that particular binding sites are enriched at TAD boundaries. It has been established that the zinc finger protein, CTCF, is associated with TAD boundaries and is an architectural protein responsible for the organization of the genome within the nucleus [@ghirlando2016ctcf; @ong2014ctcf]. CTCF along with other transcription factors, including SMC1, SMC3, and RAD21 have been shown to colocalize together near TAD boundaries, leading to the recruitment of a cohesin ring [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Other binding proteins have been shown to mediate enhancer-promoter interactions, including the transcription factors YY1 and ZNF143, thereby influencing the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [rao20143d]. However, the growing number of annotated genomic features and the methods used to identify the ones assocaited with TAD boundaries still leaves much to be desired.

Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. In doing so, many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced number of boundaries vs. nonboundary regions. This is because, when constructing the data for modeling, a common methodology used to determine the class of the response variable is to bin the linear genome and then create a method used to determine if each bin "includes" a TAD boundary or not. Given the sparcity of TADs throughout the genome though, this will inevitably lead to severly imbalanced classes.

@@

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exist several algorithmic or data level solutions. Algorithmic level solutions involve modifyng existing classifiers in order to curtail the bias induced by imbalanced classes [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Data level solutions, on the other hand, mostly constitute as resampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. For the purposes of this study, we focus our attention toward data level solutions so that all results may be classifier-independent, and, thus, more easily comparable.

Another underexplored aspect of modeling the association between TAD boundaries and genomic annotations is the various types of relationships between them. Typical approaches include overlap counts or percents [@mourad2016computational; @hong2017computational; @huang2015predicting]. Counts refer to the number of times a genomic annotation directly overlaps with a TAD boundary region, e.g., a TAD boundary may overlap with several ChIP-seq peaks for a given transcription factor. Percents refers to the percent of overlap of a genomic annotation with a TAD boundary region, e. g., a 60% of a TAD boundary may overlap genomic region annotated as heterochromatin.

While direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving force of genome regulation [??? REFS]. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. The use of distances between TAD boundaries and genomic annotations has largely been ignored. However, these distance features have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

Therefore, in this study, we aimed to developed an ensemble framework used to build random forest classifiers composed of different combinations of data driven parameters used to address the issues presented above. We used our framework to systematically compare model performances using a composite performance metric, MCC. Our goal was to determine a suitable combination of data re-sampling and feature engineering techniques that both corrected the inherent imbalanced classes and provided the best spatial representation linking genomic annotations to TAD boundaries. Additionally, we used our framework to directly compare the influence of Hi-C contact matrix resolution on model performances. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. Noninformative features were removed using a feature reduction algorithm and random forest classifications were performed to assess model performance, and investigate which genomic annotations were most associated with the TAD boundaries for a given system from the complete ensemble.

@@

Replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1). The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. For each of the 4 data resolutions, a total of 8008, 4869, 2449, and 1075 unique TADs were extracted from the GM12878 cell line, and a total of 5828, 4097, 2181, and 986 unique TADs were extracted from the K562 cell line. The TADs were distributed throughput the autosomal chromosomes, with larger chromosomes containing more TADs (Supplementary Figure 14; Supplementary Table 8). The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained.

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were performed to compare model performances between re-sampling techniques and predictor types. <!--To reduce bias due to random dataset generation, 10-fold cross validation was used to tune the number of features to consider at each node and the number of ensemble trees to aggregate-->. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was validated on the test set.

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. Additional performance metrics were reported in the supplementary materials.

@@

An ensemble framework was established for the purposes of systematically creating a variety of models that each included different sets of parameters described above (Figure 2). Each system in the framework is assumed to be composed of a model constructed from a particular data resolution, with a particular predictor type for the feature space. The data sets used for modeling were composed of the full set of genomic bins that either overlapped with a flanked TAD boundary or did not. These two classes made up the minority and majority sets respectively. The data was then split into a 7:3 ratio of training to testing sets. Each training and testing set was composed of a similar majority to minority class ratio as the full data set. For the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison. Furthermore, this process was repeated, and performances were aggregated across each autosomal chromosome. Model results for Chromosome 9 was omitted from the analysis, for both cell lines, due to the sparcity of the contact matrix at the 10 kb resolution for the K562 cell line.

@@

We compared the performances of random forest classification models implemented on TAD boundaries called using the arrowhead algorithm [@durand2016juicebox] from Hi-C data for 2 cell lines $L = \{\textit{GM12878}, \textit{K562}\}$, at four different resolutions $R = \{\textit{10 kb}, \textit{25 kb}, \textit{50 kb}, \textit{100 kb\}}$. [??? Bad spacing, for the whole paragraph. Why you are not using plain and simple Markdown?] For each classifier, we considered 3 separate predictor types as the feature space including $P = \{\textit{OC}, \textit{OP}, \textit{Distance}\}$. To assess the severity of class imbalance 4 re-sampling techniques were implemented prior to modelling, including $B = \{\textit{None}, \textit{RUS}, \textit{ROS}, \textit{SMOTE}\}$. Each combination of the above parameters was repeated 21 times (number of autosomal chromosomes, excluding chromosome 9), $C = \{ 1, 2, ..., 22 \}$. A system produced from the ensemble framework was defined as: $S = \{\textit{c}, \textit{r}, \textit{p}, \textit{b}\}$, where $l \in L$, $r \in R$, $p \in P$, $b \in B$, and $c \in C$. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2016 (= 2 \times 4 \times 3 \times 4 \times 21)$ separate models.

@@

Results showed that the imbalance ratios were well below acceptable thresholds for both cell lines, and across the 4 resolutions (Supplementary Table 9 ???SDs will come with more chromosomes). The low imbalance ratios indicated a wide disparity between the number of genomic bins that overlap with a flanked TAD boundary and the much larger number that did not. While resolution certainly played a role in determining the number of observations included in datasets used in downstream modeling, as it determines the width of genomic bins and larger widths (lower resolutions) created less bins, we found that there was very little variation in imbalance ratios across resolution (Supplementary Figure 16 ???variation will come with additional chrs).

@@

In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built data with fully imbalanced classes exhibited poor MCC across all resolutions (Figure 5A). We then gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Interestingly, we found that MCCs were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). When including distance-type predictors, models were found to exhibit slightly better MCCs than the other two types (Figure 5C). 
Importantly, when validating these 3 models built on fully imbalanced datasets, the MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Furthermore, we found that evaluating models using conventional metrics, such as AUC or accuracy, did not depict the true severity in performance (Supplementary Table 3). Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were also unable to be reported due to the models inability to produce any true positive cases.

@@

When models were built on randomly oversampled (ROS) data, we found a substantial increase in MCC for the GM12878 cell line (Figure 5 A-C). The biggest improvement in performance was seen at the 10 kb resolution in each of the 3 predictor types, with a downward trend as resolutions decreased. Although performances were low at lower resolutions, MCC values were able to be obtained, contrary to models built on completely unbalanced data as seen before. Distance-type predictors were found to offer the highest MCCs for ROS data, across all resolutions. Similar results were obtained when applying ROS to the K562 cell line (Supplementary Figure 10). Here, a much sharper negative trend was featured for MCCs as resolution decreased, than in the GM12878 cell line.

Further improvement in performance was found when models were implemented on both randomly undersampled (RUS) and SMOTE resampled data, for all 3 predictor types (Figure 5). Although the differences in performance between the two techniques were marginal, SMOTE resampled data was found to yield the best MCCs among all classes of models developed and compared in the ensemble framework. As with ROS, distance type predictors yielded the highest MCC values across resolutions, when applied to either RUS or SMOTE. SMOTE re-sampled data, with the inclusion of distance-type predictors was also found to be the best set of characteristics when applied to the K562 cell line (Supplementary Figure 10). Moreover, we found that there was a degredation in model performance as data resolution decreased. This was much more pronounced in the K562 cell line. On the GM12878 cell line, there were only marginal differences in performance between 10 kb and 25 kb, followed by a steep decline at 50 kb. Then, the MCCs showed a momentary increase in performance at 100 kb, the lowest resolution considered in the study. Performances from the K562 cell line degraded much faster as resolution decreased, starting with highest performance at 10 kb and ending with the worst performance at 100 kb. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on distance-type predictors and incorporates both random under- and over-sampling.

@@

In order to assess the predictive capabilities of specific types of annotations, we further compared the performances of annotation-specific models. Using the optimal set of parameters, SMOTE re-sampling with distance-type predictors, random forest classifiers built on sets of histone modifications, chromatin state segmentations, and TFBS were compared to a model that included all three types together. For the GM12878 cell line, models built simply on TFBS were found to perform consistently just as well as models that included all 3 sets of annotations, across resolution (Figure 6). Similar results were found when evaluating the K562 cell line (supplementary Figure 11). For both resolutions, models built on only histone modifications were found to have the MCC, across resolutions, followed by models built on chromatin state segmentations. This suggests that TAD boundary formation is primarily carried out by transcription factor binding sites.

@@

As an external method of validation, we also investigated the predictive importance of the annotations that our top performing models were built on. We found that specific transcription factor binding sites (TFBS), including SMC3ab, ZNF143, CTCF, and RAD21, were given consistently high predictive importance in the models built on the GM12878 cell line, across all resolutions (Figure 8). These are known factors associated with the hypothesized loop extrusion model [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self] and were found to yield consistently high predictive importance on the K562 cell line as well, with similar profiles across resolutions (Supplementary Figure 15).

@@

When evaluating the 10 kb data resolution, on GM12878, the distance profiles (on a $log_{2}$ scale) were found to be smaller for top predictive TFBS annotations in relation to genomic bins that overlapped with flanked TAD boundaries, when compared to bins that did not overlap with a flanked TAD boundary (Figure 7). At the same resolution and cell line, direct comparisons of quantiles obtained from the distribution of distances for the transcription factors SMC3a, ZNF143, CTCF, and RAD21, led to the the conclusion that these TFBS were more often located near TAD boundaries on the linear genome, as can be seen from the shifted quantile differences (Figure 7). Similar distance profiles for the same TFBS were obtained when assessed at lower resolutions, across both cell lines (Supplementary Figures 12-17 A-D).

>>>>

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. Likewise, emerging research has shown that particular binding sites are enriched at TAD boundaries. It has been established that the zinc finger protein, CTCF, is associated with TAD boundaries and is an architectural protein responsible for the organization of the genome within the nucleus [@ghirlando2016ctcf; @ong2014ctcf]. CTCF along with other transcription factors, including SMC1, SMC3, and RAD21 have been shown to colocalize together near TAD boundaries, leading to the recruitment of a cohesin ring [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Other binding proteins have been shown to mediate enhancer-promoter interactions, including the transcription factors YY1 and ZNF143, thereby influencing the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [@rao20143d]. However, the growing number of annotated genomic features and the methods used to identify the ones assocaited with TAD boundaries still leaves much to be desired.

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. In doing so, many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced number of boundaries vs. nonboundary regions. This is because, when constructing the data for modeling, a common methodology used to determine the class of the response variable is to bin the linear genome and then create a method used to determine if each bin "includes" a TAD boundary or not. Given the sparcity of TADs throughout the genome though, this will inevitably lead to severly imbalanced classes.

@@

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exists several algorithmic or data level solutions. Algorithmic level solutions involve modifyng existing classifiers in order to curtail the bias induced by imbalanced classes [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Data level solutions, on the other hand, mostly constitute as re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. For the purposes of this study, we focus our attention toward data level solutions so that all results may be classifier-independent, and, thus, more easily comparable.

Another underexplored aspect of modeling the association between TAD boundaries and genomic annotations is determining a suitable representation of the relationship between them. Typical approaches include overlap counts or percents [@mourad2016computational; @hong2017computational; @huang2015predicting]. Counts refer to the number of times a genomic annotation directly overlaps with a TAD boundary region, e.g., a TAD boundary may overlap with several ChIP-seq peaks for a given transcription factor. Percents refers to the percent of overlap between a genomic annotation and a TAD boundary region, e. g., 60% of a TAD boundary region may overlap with a genomic region annotated as heterochromatin.

While direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving force of genome regulation [??? REFS]. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. On the other hand, the use of distances between TAD boundaries and genomic annotations has largely been ignored. However, these distance features have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

Therefore, in this study, we aimed to develop an ensemble framework used to build random forest classifiers composed of different combinations of data driven parameters used to address the issues presented above. We used our framework to systematically compare model performances using a composite performance metric, MCC. Our goal was to determine a suitable combination of data re-sampling and feature engineering techniques that both corrected the inherent imbalanced classes and provided the best spatial representation linking genomic annotations to TAD boundaries. Additionally, we used our framework to directly compare the influence of Hi-C contact matrix resolution on model performances. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. Noninformative features were removed using a feature reduction algorithm and random forest classifications were built to assess model performance, and investigate which genomic annotations were most predictive of TAD boundaries for a given system from the complete ensemble.

@@

Replicate Hi-C data for the GM12878 and K562 cell lines were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1). The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. For each of the 4 data resolutions, a total of 8008, 4869, 2449, and 1075 unique TADs were extracted from the GM12878 cell line, and a total of 5828, 4097, 2181, and 986 unique TADs were extracted from the K562 cell line. The TADs were distributed throughput the autosomal chromosomes, with larger chromosomes containing more TADs (Supplementary Figure 14; Supplementary Table 8). The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained. TAD boundaries called on Chromosome 9 were omitted from all downstream analysis, for both cell lines, due to the sparcity of the contact matrix at the 10 kb resolution for the K562 cell line.

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were built to compare model performances between re-sampling techniques and predictor types. <!--To reduce bias due to random dataset generation, 10-fold cross validation was used to tune the number of features to consider at each node and the number of ensemble trees to aggregate.--> The default number of features to consider at each node of the random forest algorithm was set as $\sqrt{p}$, where $p$ refers to the number of predictors included in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was validated on the test set.

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. Additional performance metrics were reported in the supplementary materials.

@@

An ensemble framework was then established for the purposes of systematically creating a variety of models that each included different sets of parameters described above (Figure 2). Each system in the framework is assumed to be composed of a model constructed from a particular cell line and data resolution, with a particular predictor type for the feature space. The data sets used for modeling were composed of the full set of genomic bins that either overlapped with a flanked TAD boundary or did not. The data was then split into a 7:3 ratio of training to testing sets. For the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison. Furthermore, this process was repeated, and performances were aggregated across each autosomal chromosome (omitting chromosome 9) using means and standard deviations.

@@

We compared the performances of random forest classification models implemented on TAD boundaries called using the arrowhead algorithm [@durand2016juicebox] from replicate Hi-C data for 2 cell lines $L = \{\textit{GM12878}, \textit{K562}\}$, at four different resolutions $R = \{\textit{10 kb}, \textit{25 kb}, \textit{50 kb}, \textit{100 kb\}}$. [??? Bad spacing, for the whole paragraph. Why you are not using plain and simple Markdown?] For each classifier, we considered 3 separate predictor types as the feature space including $P = \{\textit{OC}, \textit{OP}, \textit{Distance}\}$. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{\textit{None}, \textit{RUS}, \textit{ROS}, \textit{SMOTE}\}$. Each combination of the above parameters was repeated 21 times (number of autosomal chromosomes, excluding chromosome 9), $C = \{ 1, 2, ..., 22 \}$. A system produced from the ensemble framework was defined as: $S = \{\textit{c}, \textit{r}, \textit{p}, \textit{b}\}$, where $l \in L$, $r \in R$, $p \in P$, $b \in B$, and $c \in C$. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2016 (= 2 \times 4 \times 3 \times 4 \times 21)$ separate models.

@@

Results showed that the imbalance ratios were well below acceptable thresholds for both cell lines, and across the 4 resolutions (Supplementary Table 9 ???SDs will come with more chromosomes). The low imbalance ratios indicated a wide disparity between the number of genomic bins that overlaped with a flanked TAD boundary and the much larger number that did not. While resolution certainly played a role in determining the number of observations included in datasets used in downstream modeling, as it determines the width of genomic bins and larger widths (lower resolutions) created less bins, we found that there was very little variation in imbalance ratios across resolution (Supplementary Figure 16 ???variation will come with additional chrs).

@@

In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 5A). We then gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Interestingly, we found that MCCs were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). When including distance-type predictors, models were found to exhibit slightly better MCCs than the other two types (Figure 5C). Importantly, when validating these 3 models built on fully imbalanced datasets, the MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Furthermore, we found that evaluating models using conventional metrics, such as AUC or accuracy, did not depict the true severity in performance (Supplementary Table 3). Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were also unable to be reported due to the models inability to produce any true positive cases.

@@

When models were built on randomly oversampled (ROS) data, we found a substantial increase in MCC for the GM12878 cell line (Figure 5 A-C). The biggest improvement in performance was seen at the 10 kb resolution in each of the 3 predictor types, with a downward trend as resolutions decreased. Although performances were low at lower resolutions, MCC values were able to be obtained, contrary to models built on completely unbalanced data as seen before. Distance-type predictors were found to offer the highest MCCs for ROS data, across all resolutions. Similar results were obtained when applying ROS to the K562 cell line (Supplementary Figure 10). For K562, a much sharper negative trend was featured for MCCs as resolution decreased, than in the GM12878 cell line.

Further improvement in performance was found when models were implemented on both randomly undersampled (RUS) and SMOTE resampled data, for all 3 predictor types and across resolutions (Figure 5). SMOTE resampled data was found to yield the best MCCs among all classes of models developed and compared in the ensemble framework. Here, distance type predictors were found to again yield the highest MCC values across resolutions. SMOTE re-sampled data, with the inclusion of distance-type predictors was also found to be the best set of characteristics when applied to the K562 cell line (Supplementary Figure 10). Moreover, for these sets of data characteristics, we again found that there was a degredation in model performance as data resolution decreased. This was much more pronounced in the K562 cell line. On the GM12878 cell line, there were only marginal differences in performance between 10 kb and 25 kb, followed by a steep decline at 50 kb. Then, the MCCs showed a momentary increase in performance at 100 kb, the lowest resolution considered in the study. Performances from the K562 cell line degraded much faster as resolution decreased, starting with highest performance at 10 kb and ending with the worst performance at 100 kb. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on distance-type predictors and incorporates both random under- and over-sampling. Furthermore, models built on TAD boundaries called at higher resolutions exhibited better performances.

@@

In order to assess the predictive capabilities of specific types of annotations, we further compared the performances of annotation-specific models. Using the optimal set of characteristics obtained above (SMOTE re-sampling with distance-type predictors), random forest classifiers built on sets of cell line specific histone modifications, chromatin state segmentations, and transcription factor binding sites (TFBS) were compared to a model that included all three types together. For GM12878, models built simply on TFBS were found to perform consistently just as well as models that included all 3 sets of annotations, across resolution (Figure 6). Similar results were found when evaluating the K562 cell line (supplementary Figure 11). For both resolutions, models built on only histone modifications were found to have the lowest MCC, across resolutions, followed by models built on chromatin state segmentations. This suggests that TAD boundary formation is primarily carried out by transcription factor binding sites.

@@

As an external method of validation, we also investigated the predictive importance of the annotations that our top performing models were built on. We found that specific TFBS, including SMC3ab, ZNF143, CTCF, and RAD21, were given consistently high predictive importance in the models built on the GM12878 cell line, across all resolutions (Figure 8). These are known factors associated with the hypothesized loop extrusion model [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self] and were found to yield consistently high predictive importance on the K562 cell line as well, with similar profiles across resolutions (Supplementary Figure 15). [???mention additional tfbs when full results ready]

@@

<!--
When evaluating the 10 kb data resolution, on GM12878, the distance profiles (on a $log_{2}$ scale) were found to be smaller for top predictive TFBS annotations in relation to genomic bins that overlapped with flanked TAD boundaries, when compared to bins that did not overlap with a flanked TAD boundary (Figure 7). At the same resolution and cell line, direct comparisons of quantiles obtained from the distribution of distances for the transcription factors SMC3a, ZNF143, CTCF, and RAD21, led to the the conclusion that these TFBS were more often located near TAD boundaries on the linear genome, as can be seen from the shifted quantile differences (Figure 7). Similar distance profiles for the same TFBS were obtained when assessed at lower resolutions, across both cell lines (Supplementary Figures 12-17 A-D).
-->
